---
title: Problem set 4
date: 2025-10-05
format: html
---

In the next problem set, we plan to explore the relationship between COVID-19 death rates and vaccination rates across US states by visually examining their correlation. This analysis will involve gathering COVID-19 related data from the CDC's API and then extensively processing it to merge the various datasets. Since the population sizes of states vary significantly, we will focus on comparing rates rather than absolute numbers. To facilitate this, we will also source population data from the US Census to accurately calculate these rates. 

In this problem set we will learn how to extract and wrangle data from the data US Census and CDC APIs.

1. Get an API key from the US Census at <https://api.census.gov/data/key_signup.html>. You can't share this public key. But your code has to run on a TFs computer. Assume the TF will have a file in their working directory named `census-key.R` with the following one line of code:

```
census_key <- "A_CENSUS_KEY_THAT_WORKS"
```

Write a first line of code for your problem set that defines `census_key` by running the code in the file `census-key.R`.

```{r}
census_key <- source("census-key.R") # gets census_key from census-key.R file
```

2. The [US Census API User Guide](https://www.census.gov/content/dam/Census/data/developers/api-user-guide/api-user-guide.pdf)
provides details on how to leverage this valuable resource. We are interested in vintage population estimates for years 2021 and 2022. From the documentation we find that the _endpoint_ is:


```{r}
url <- "https://api.census.gov/data/2021/pep/population"
```

Use the **httr2** package to construct the following GET request.

```
https://api.census.gov/data/2021/pep/population?get=POP_2020,POP_2021,NAME&for=state:*&key=YOURKEYHERE
```

Create an object called `request` of class `httr2_request` with this URL as an endpoint.
Hint: Print out `request` to check that the URL matches what we want.
```{r}
#| message: false
#| warning: false
library(httr2)
request <- request(url) |> # creates HTTP GET request
  req_url_query(
    get = "POP_2020,POP_2021,NAME",
    `for`="state:*",
    key = census_key$value) # use key to access API
```

3. Make a request to the US Census API using the `request` object. Save the response to and object named `response`. Check the response status of your request and make sure it was successful. You can learn about _status codes_ [here](https://en.wikipedia.org/wiki/List_of_HTTP_status_codes).

```{r}
response <- request |> req_perform() # send request to API web server, US Census
response$status_code # successful code is 200
```

4. Use a function from the **httr2** package to determine the content type of your response.

```{r}
resp_content_type(response)
```

5. Use just one line of code and one function to extract the data into a matrix. 
Hints: 1) Use the `resp_body_json` function. 2) The first row of the matrix will be the variable names and this OK as we will fix in the next exercise.

```{r}
population <- resp_body_json(response,simplifyVector = TRUE) # converts each list into a column and the elements of the list into the cells of the matrix
```

6. Examine the `population` matrix you just created. Notice that 1) it is not tidy, 2) the column types are not what we want, and 3) the first row is a header. Convert `population` to a tidy dataset. Remove the state ID column and change the name of the column with state names to `state_name`. Add a column with state abbreviations called `state`. Make sure you assign the abbreviations for DC and PR correctly. Hint: Use the **janitor** package to make the first row the header. 

```{r}
#| message: false
#| warning: false
library(tidyverse)
library(janitor)
population <- population |> 
  row_to_names(row_number = 1) |> # use janitor row to names function to make column names the header and not a row
  as_tibble() |> # convert to tibble
  select(-state) |>  # remove state column
  rename(state_name = NAME) |> # rename state column to state_name
  pivot_longer(starts_with("POP_"), names_to = "year", values_to = "population") |> # increase number of rows and decrease number of columns so there are two rows per state
  mutate(year = str_remove(year, "POP_")) |> # remove POP_ from year
  mutate(state = case_when(state_name == "Puerto Rico"~"PR", 
                           state_name == "District of Columbia" ~ "DC",
  TRUE ~ state.abb[match(state_name, state.name)]), # add state abbreviations using state.abb variable
  # use case_when to add abbreviations for DC and PR
  population = as.numeric(population),
  year = as.numeric(year)) # parse all relevant columns to numeric
```

7. As a check, make a barplot of states' 2021 and 2022 populations. Show the state names in the y-axis ordered by population size.  Hint: You will need to use `reorder` and use `facet_wrap`.

```{r}
population |> 
  ggplot(aes(population, reorder(state_name, population))) + # reorder state levels by population size for plotting
  geom_col() + # use geom_col to plot barplot so state names can appear alongside their respective population size
  facet_wrap(~year) + # facet by year
  labs(x = "Population size", y = "State name", title = "State Populations in 2020 and 2021")
```


8. The following URL:

```{r}
url <- "https://github.com/datasciencelabs/2025/raw/refs/heads/main/data/regions.json"
```

points to a JSON file that lists the states in the 10 Public Health Service (PHS) defined by CDC. We want to add these regions to the `population` dataset. To facilitate this create a data frame called `regions` that has two columns `state_name`, `region`, `region_name`. One of the regions has a long name. Change it to something shorter.

```{r}
#| message: false
#| warning: false
library(jsonlite)
library(purrr)
url <- "https://github.com/datasciencelabs/2025/raw/refs/heads/main/data/regions.json"

regions <- url |>
  fromJSON() |>  # use jsonlite JSON parser; this automatically downloads and parses the JSON into a data frame
  rename(state_name = states) |>
  unnest(c(state_name, region)) |> # display the elements in the nested lists
  mutate(region_name = case_when(region_name == "New York and New Jersey, Puerto Rico, Virgin Islands" ~ "Other",
         TRUE ~ region_name)) |> # keep all other region names unchanged 
  as.data.frame() 
```

9. Add a region and region name columns to the `population` data frame.

```{r}
population <- left_join(population, regions, by = "state_name") # keep all rows in population and add the rows from region that overlap with population data frame 
  
```


10. From reading <https://data.cdc.gov/> we learn the endpoint `https://data.cdc.gov/resource/pwn4-m3yp.json` provides state level data from SARS-COV2 cases. Use the **httr2** tools you have learned to download this into a data frame. Is all the data there? If not, comment on why.

We see exactly 1,000 rows. We should be seeing over $52 \times 3$ rows per state.
```{r}
api <- "https://data.cdc.gov/resource/pwn4-m3yp.json"

cases_raw <- request(api) |> # get HTTP GET request
  req_perform() |> # send request to CDC website
  resp_body_string() |> # extract data and returns in plain text
  fromJSON() # convert into data frame
```


11. The reason you see exactly 1,000 rows is because CDC has a default limit. You can change this limit by adding `$limit=10000000000` to the request. Rewrite the previous request to ensure that you receive all the data. Then wrangle the resulting data frame to produce a data frame with columns `state`, `date` (should be the end date) and `cases`. Make sure the cases are numeric and the dates are in `Date` ISO-8601 format.

```{r}
api <- "https://data.cdc.gov/resource/pwn4-m3yp.json"

library(lubridate)
cases_raw <- request(api) |> # form request
  req_url_query(`$limit` = 10000000000) |> # increase rate limit so large data can get extracted
  req_perform() |> # send request to CDC website
  resp_body_string() |> # extract data from server and returns in plain text form
  fromJSON() |> # convert into dataframe
  drop_na(state,end_date,new_cases)|> # drop all missing values
  rename(date = end_date, 
         cases = new_cases) |>
  mutate(cases = as.numeric(cases), date = as.Date(date)) |> # convert  cases to be numeric and reassign to cases and convert date to ISO-8601 format (YYYY-MM-DD)
  select(state, date, cases)
```


12. For 2020 and 2021, make a time series plot of cases per 100,000 versus time for each state. Stratify the plot by region name. Make sure to label you graph appropriately. 

```{r}
cases_raw |>
  filter(str_starts(date, "2020") | str_starts(date, "2021")) |> # only include years from 2020 and 2021
  left_join(population, by = "state") |> # join population to cases_raw so region_name can be used for faceting 
  drop_na(region_name, state,date,population,cases) |> # drop all missing values after joining the two datasets
  ggplot(aes(date, (10^5)*cases/population, col = state)) + # color each line by state
  geom_line() +
  labs(x="Date", y = "Cases per 100,000", title = "Total COVID-19 Cases by Region and State in 2020-2021") + facet_wrap(~region_name, ncol=2) # 10 regions so have 5 regions in one column, 5 regions in the other
```

13. The dates in the `cases` dataset are stored as character strings. Use the **lubridate** package to properly parse the `date` column, then create a summary table showing the total COVID-19 cases by month and year for 2020 and 2021. The table should have columns for year, month (as month name), and total cases across all states. Order by year and month. Use the **knitr** package and `kable()` function to display the results as a formatted table.

```{r}
library(knitr)
cases_raw |>
  filter(str_starts(date, "2020") | str_starts(date, "2021"))|> # keep cases from 2020 and 2021
  mutate(year = year(date), # extracts year from date variable
         month = month(date, label=TRUE)) |> # extracts month from date and then convert to month name using label=TRUE
  select(year, month, cases) |>
  group_by(year, month) |>
  summarize(total_cases = sum(cases)) |>
  kable()
```

14. The following URL provides additional COVID-19 data from the CDC in JSON format:

```{r}
deaths_url <- "https://data.cdc.gov/resource/9bhg-hcku.json"
```

Use **httr2** to download COVID-19 death data from this endpoint. Make sure to remove the default limit to get all available data. Create a clean dataset called `deaths` with columns `state`, `date`, and `deaths` (renamed from the original column name). Ensure dates are in proper Date format and deaths are numeric.
```{r}
deaths <- request(deaths_url) |> # creates HTTP GET request
  req_url_query(`$limit`= 10000000000) |> # removes default limit by setting limit to a large number
  req_perform() |> # send request to API web server, CDC
  resp_body_string() |> # returns data in plain text form
  fromJSON() |> # returns data as a dataframe
  drop_na(covid_19_deaths, end_date) |>
  filter(state != "United States") |> # filter out U.S. from state variable
  rename(deaths = covid_19_deaths,
         date = end_date) |>
  mutate(deaths = as.numeric(deaths),
         date = as.Date(date)) |>
  select(state, date, deaths)
```

15. Using the `deaths` dataset you created, make a bar plot showing the total COVID-19 deaths by state. Show only the top 10 states with the highest death counts. Order the bars from highest to lowest and use appropriate labels and title.

```{r}
deaths |>
  group_by(state) |>
  summarize(tot_deaths = sum(deaths),
            .groups = "drop") |> # compute total deaths per state over the years
  arrange(desc(tot_deaths)) |> # arrange deaths from highest to lowest
  head(10) |> # show only the top 10 states with the highest death counts
  ggplot(aes(reorder(state, tot_deaths, decreasing=TRUE), tot_deaths)) + 
  geom_col() +
  labs(x="States", 
       y="Total COVID-19 Deaths",
       title="Top 10 States with Highest COVID-19 Deaths in 2020-2023")
```